Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-20 rnni_convexity.tex
up to explanation csort

In this section we aim to investigate properties of paths between \emph{caterpillar trees}, each internal node of which is adjacent to at least one leaf.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called a \emph{caterpillar path}.
If $p$ is the shortest pass among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.

Let us now introduce a representation of caterpillar trees as lists of taxa.
This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
\todo{I don't follow these two sentences.}
For a caterpillar tree with a cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
In this list representing a caterpillar tree the taxa are ordered non-decreasing, according to the ranks of their parents.
Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
That is why we order them in the list such that
\todo{Are we assuming taxa is ordered?}
$x_1 < x_2$
to make the list representation of caterpillar trees unique.
We will call a pair of two taxa that share their parent a \emph{cherry}.

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$.
This path $p$ is built stepwise by adding a new list to the end of $p$.
We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.

>>>>

In this section we introduce an algorithm to compute paths between \emph{caterpillar trees}, which are trees where each internal node is adjacent to at least one leaf.
As a result caterpillar trees have only one \emph{cherry}, which is a pair of taxa that share their parent.
For introducing this algorithm we need the following definitions.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called a \emph{caterpillar path}.
If $p$ is the shortest path among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.

Let us now introduce a representation of caterpillar trees as lists of taxa, which is a simplification of the cluster representation for caterpillar trees and works as follows.
In the list representing a caterpillar tree its taxa are ordered non-decreasing, according to the ranks of their parents.
Note that the only two elements in such a list that have parents of equal rank are the first two elements $l_1$ and $l_2$.
As we assume to have taxon set $X = \{1, \ldots, n\}$, we assume that they are order in the list such that $l_1 < l_2$, according to their ordering as natural numbers.
This ensures that the list representation of caterpillar trees is unique.

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$ as introduced above.
The output path $p$ is built stepwise by adding a new list to the end of $p$.
We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.
